Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2011-02-08 mnras/mnras.tex
Updating figures for updated tests 3 and 7.  Adding Mach number plot of test 7.
Minor changes to text.  Adding PDFs to repo but not paper yet.
Adding PDF figures into the manuscript along with their descriptions in the text.

\begin{figure}
  %\epsscale{0.8}
  \plotone{f6.eps}
  \caption{\label{fig:test1_HI} Test 1 (\hii region expansion
    with a monochromatic spectrum of 13.6 eV). Slice of neutral
    fraction at the origin at 500 Myr.}
  %\epsscale{1.0}
\end{figure}

@@

In Fig. \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionisation fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionisation front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius [equation (\ref{eqn:rstr})].  The ionisation
front has a width of $\sim0.7$ kpc, which is in agreement with the
inherent thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a
13.6 eV mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii.  This occurs because the initial ionised fraction and
temperature is set to $1.2 \times 10^{-3}$ and $10^4$ K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Fig. \ref{fig:test1_ifront}, the
ionisation front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
[equation (\ref{eqn:rstr})] assumes the \hii region has a constant
ionised fraction.  The evolution of the ionised fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionisation front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  In Fig. \ref{fig:test1_HI}, we show a slice of the neutral
fraction through the origin.  Other than the ray splitting artifacts
that generate the plateaus at 1.5 and 3 kpc, one sees spherical
symmetry without any noise in our solution.

@@

@@

In Fig. \ref{fig:test2_1a}, we show the radially averaged neutral
and ionised fraction at $t = $ 10, 30, 100, and 500 Myr, and the total
neutral fraction of the domain in Fig. \ref{fig:test2_1b}.  Compared
with Test 1, the ionisation front is thicker, as expected with the
harder spectrum.  The total neutral fraction decreases to 0.67 over
4$t_{\rm rec} = 500$ Myr, which is in agreement with the analytical
expectation and other codes in RT06.  In Fig. \ref{fig:test2_2}, we
show slices of neutral fraction and temperature through the origin at
$t = $ 10 and 100 Myr.  Here one sees the spherically symmetric \hii
regions and a smooth temperature transition to the neutral ambient
medium.  In Fig. \ref{fig:test2_3}, we show the ratio of the
ionisation front radius $r_{\rm IF}$ in our simulation and $R_s$.
Before $1.5t_{\rm rec}$, $r_{\rm IF}$ lags behind $R_s$, initially by
10\% and then increases to $R_s$; however afterwards, this ratio
asymptotes to a solution that is 4\% greater than $R_s$.  This
behavior is approximately the median result in RT06, where this ratio
varies between 1 and 1.1, and the early evolution of $r_{\rm IF}$ is
under-predicted by almost all of the codes.  If we use one energy
group with the mean energy (29.6 eV) of a $T=10^5$ K blackbody, we
find that $r_{\rm IF}/R_s$ = 1.08, which is representative of the
codes in the upper range of RT06.

@@

\begin{figure}
  \plottwo{f11a}{f11b}
  \caption{\label{fig:test3_1} Test 3. (I-front trapping in a dense
    clump and shadowing).  Line cut from the point source through the
    middle of the dense clump at $t = 1, 3, 5, 15$ Myr.  of the
    average neutral fraction (top) and temperature (bottom) of the
    clump.}
\end{figure}

\begin{figure}
  %\epsscale{0.85}
  \plotone{f12}
  \caption{\label{fig:test3_2} Test 3. (I-front trapping in a dense
    clump and shadowing).  Evolution of the average ionised fraction
    (top) and temperature (bottom) of the overdense clump.}
  %\epsscale{1}
\end{figure}

@@

The problem for this test is contained in a 6.6 kpc box with an
ambient medium of $n_H = 2 \times 10^{-4}\;\cubecm$ and $T = 8000$ K.
The clump is in pressure equilibrium with $n_H = 0.04\; \cubecm$ and
$T = 40$ K.  It has a radius of $r_c = 0.8$ kpc and is centred at
$(x,y,z) = (5, 3.3, 3.3)$ kpc.  In RT06, the test considered a plane
parallel radiation field with a flux $F_0 = 10^6$ ph s$^{-1}$
cm$^{-2}$ originating from the $y=0$ plane.  Our code can only
consider point sources, so we use a single radiation source located in
the centre of the $y=0$ boundary.  The luminosity of $\dot{N}_\gamma =
3 \times 10^{51}$ ph s$^{-1}$ corresponds to the same flux $F_0$ at 5
kpc.  The location where the ionisation front trapping can be
calculated analytically \citep{Shapiro04}, and with these parameters,
it should halt at approximately the centre of the clump.  We use the
same four energy groups as in Test 2.

In Fig. \ref{fig:test3_1}, we show neutral fraction and temperature
of a one-dimensional cut through the centre of the dense clump at $z =
0.5$ at $t = 1, 3, 5, 15$ Myr.  The ionisation front is halted at a
little more than halfway through the clump, which is consistent with
the analytical expectation.  The hardness of the $T = 10^5$ K
blackbody spectrum allows the gas outside of the ionisation front to
be photo-heated.  Where the gas is ionised, the temperature is between
10,000 and 20,000 K, but decreases sharply with the ionised fraction.
Fig. \ref{fig:test3_2} depicts the average ionised fraction and
temperature inside the dense clump, which both gradually increase as
the ionisation front propagates through the overdensity.  Our results
are consistent with RT06.  Finally we show slices of neutral fraction
and temperature in the $z = 0.5$ plane in Fig. \ref{fig:test3_3}.
The neutral fraction slices prominently show the sharp shadows created
by the clump and demonstrates the non-diffusivity behavior of ray
tracing.  The discretisation of the sphere creates one neutral cell on
the left side of the sphere.  This inherent artifact to the initial
setup carries through the calculation.  We did not smooth the clump
surface like in some of the RT06 codes, in order to remove this
artifact.  It is seen in the neutral fraction and temperature states
at all times and is not a caused by our ray tracing algorithm.

@@

\begin{figure}
  %\epsscale{0.75}
  \plotone{f14}
  \caption{\label{fig:test4_1} Test 4. (Multiple cosmological
    sources).  Evolution of the mass- (dashed) and volume-averaged
    (solid) ionised fraction.}
  %\epsscale{1}
\end{figure}

@@

We show the growth of the \hii regions by computing the
mass-averaged $x_m$ and volume-averaged $x_v$ ionised fraction in
Fig. \ref{fig:test4_1}.  Initially $x_m$ is larger than $x_v$, and
at $t \sim 170$ kyr, the $x_v$ becomes larger.  This is indicative of
inside-out reionisation \citep[e.g.][]{Gnedin00, Miralda00,
  Sokasian03}, where the dense regions around haloes are ionised first,
then the voids are ionised last.  At the end of the simulation, 65\%
of the simulation is ionised.  However by visual inspection in the
slices of electron fraction (Fig. \ref{fig:test4_2}), there appears to
be very good agreement with C$^2$-ray and FTTE.  By first glance, our
result appears to be different than the RT06 because of the
color-mapping.  Our results are also in good agreement with the
multi-frequency version of TRAPHIC \citep[][see also for better
representations of the electron fraction slices]{Pawlik10}.  In the
slices of electron fraction and temperature, Fig. \ref{fig:test4_2},
the photo-heated regions are larger than the ionised regions by a
factor of 2--3 because of the hardness of the $T = 10^5$ K blackbody
spectrum.

@@

@@

Fig. \ref{fig:test7_1} shows cuts of density, temperature, neutral
fraction, and pressure in a line connecting the source and the clump
centre at $t = 1$, 10, and 50 Myr.  At 1 Myr, the ionisation front has
propagated through the left-most 500 pc of the clump.  This heated gas
is now over-pressurised, as seen in the pressure plot in Fig.
\ref{fig:test7_1}, and then expands into the ambient medium.  This
expansion creates a photo-evaporative flow, seen in many star forming
regions \citep[e.g. M16;][]{Hester96} as stars irradiate nearby cold,
dense overdensities.  These flows become evident in the density at 10
Myr, seen both in the line cuts and slices (Fig. \ref{fig:test7_2}).
They have temperatures up to 50,000 K.  At this time, the front has
progressed about halfway through the clump, if one inspects the
neutral fraction.  However the high energy photons have heated all but
the rear surface of the clump.  At the end of the test ($t = 50$ Myr),
only the core and its associated shadow is neutral, seen in Fig.
\ref{fig:test7_3}.  The core has been compressed by the surrounding
warm medium, thus causing the higher densities seen at $t = 50$ Myr.
The non-spherical artifacts on the inner boundary of the warm
outermost shell are caused by the initial discretisation of the
sphere, as discussed in \S\ref{sec:test3}.

@@

We thank Richard Klein for helpful comments on the manuscript.
J.~H.~W. is supported by NASA through Hubble Fellowship grant
\#120-6370 awarded by the Space Telescope Science Institute, which is
operated by the Association of Universities for Research in Astronomy,
Inc., for NASA, under contract NAS 5-26555.  Computational resources
were provided by NASA/NCCS award SMD-09-1439.  The majority of the
analysis and plots were done with \texttt{yt} \citep{yt_full_paper}.

>>>>

\begin{figure}
  \plotone{test1_pdf.eps}
  \caption{\label{fig:test1_pdf} Test 1 (\hii region expansion with a
    monochromatic spectrum of 13.6 eV). Probability distribution
    function for neutral fraction at 10 Myr (solid), 100 Myr (dashed),
    and 500 Myr (dotted).  Recombination of hydrogren at T = $10^4$ K
    causes the maximum neutral fraction to decrease from 1 to 0.75.}
\end{figure}

\begin{figure}
  \plotone{f6.eps}
  \caption{\label{fig:test1_HI} Test 1 (\hii region expansion
    with a monochromatic spectrum of 13.6 eV). Slice of neutral
    fraction at the origin at 500 Myr.}
\end{figure}

@@

In Fig. \ref{fig:test1_ifront}, we show (a) the evolution of the
neutral and ionisation fraction as a function of radius at $t = $ 10,
30, 100, and 500 Myr, and (b) the growth of the ionisation front
radius as a function of time and its ratio with the analytical
Str\"{o}mgren radius [equation (\ref{eqn:rstr})].  The ionisation
front has a width of $\sim0.7$ kpc, which is in agreement with the
inherent thickness of $\sim18 \lambda_{\rm mfp} = 0.74$ kpc, given a
13.6 eV mono-chromatic spectrum.  There are small kinks in the neutral
fraction at 1.5 and 3 kpc that corresponds to artifacts created by the
photon package splitting at these radii.  However these do not affect
the overall solution.  One difference between our results and the
codes presented in RT06 is the increasing neutral fraction outside of
the \hii region.  This occurs because the initial ionised fraction and
temperature is set to $1.2 \times 10^{-3}$ and $10^4$ K, which are not
the equilibrium values.  Over the 500 Myr in the calculation, the
neutral fraction increases to 0.2, which is close to its equilibrium
value.  In the right panel of Fig. \ref{fig:test1_ifront}, the
ionisation front radius exceeds $R_s$ by a few percent for most of the
calculation.  This difference happens because the analytical solution
[equation (\ref{eqn:rstr})] assumes the \hii region has a constant
ionised fraction.  The evolution of the ionised fraction as a function
of radius can be analytically calculated \citep[e.g.][]{Osterbrock89,
  Petkova09}, causing the ionisation front radius to be slightly
larger, increasing from 0 to 3\% in the interval 80--350 Myr.  Our
results are in excellent agreement with this more accurate analytical
solution.  We show the distribution of neutral fraction in the domain
for $t$ = 10, 100, and 500 Myr in Fig. \ref{fig:test1_pdf} that agrees
well with the results in RT06.  In Fig. \ref{fig:test1_HI}, we show a
slice of the neutral fraction through the origin.  Other than the ray
splitting artifacts that generate the plateaus at 1.5 and 3 kpc, one
sees spherical symmetry without any noise in our solution.

@@

\begin{figure}
  \plotone{test2_pdf}
  \caption{\label{fig:test2_pdf} Test 2. (\hii region expansion with a
    $T=10^5$ K blackbody spectrum).  Probability distribution function
    for ionized fraction (left) and temperature (right) at 10 Myr
    (solid), 100 Myr (dashed), and 500 Myr (dotted).}
\end{figure}

@@

In Fig. \ref{fig:test2_1a}, we show the radially averaged neutral and
ionised fraction at $t = $ 10, 30, 100, and 500 Myr, and the total
neutral fraction of the domain in Fig. \ref{fig:test2_1b}.  Compared
with Test 1, the ionisation front is thicker, as expected with the
harder spectrum.  Artifacts originating from ray splitting, similar to
Test 1, appear at $r \sim 1$ and 3 kpc as kinks in the neutral
fraction.  The total neutral fraction decreases to 0.67 over 4$t_{\rm
  rec} = 500$ Myr, which is in agreement with the analytical
expectation and other codes in RT06.  Fig. \ref{fig:test2_pdf} shows
the distribution of ionized fraction and temperature in Test 2.  The
overall trends agree with the codes presented in RT06 with the
exception of the ray splitting artifacts that appear as slight rises
in the distribution at $\log x_e \sim -1$ and $\log T \sim 3$. In
Fig. \ref{fig:test2_2}, we show slices of neutral fraction and
temperature through the origin at $t = $ 10 and 100 Myr.  Here one
sees the spherically symmetric \hii regions and a smooth temperature
transition to the neutral ambient medium.  In Fig. \ref{fig:test2_3},
we show the ratio of the ionisation front radius $r_{\rm IF}$ in our
simulation and $R_s$.  Before $1.5t_{\rm rec}$, $r_{\rm IF}$ lags
behind $R_s$, initially by 10\% and then increases to $R_s$; however
afterwards, this ratio asymptotes to a solution that is 4\% greater
than $R_s$.  This behavior is approximately the median result in RT06,
where this ratio varies between 1 and 1.1, and the early evolution of
$r_{\rm IF}$ is under-predicted by almost all of the codes.  If we use
one energy group with the mean energy (29.6 eV) of a $T=10^5$ K
blackbody, we find that $r_{\rm IF}/R_s$ = 1.08, which is
representative of the codes in the upper range of RT06.

@@

\begin{figure}
  \plotone{f11}
  \caption{\label{fig:test3_1} Test 3. (I-front trapping in a dense
    clump and shadowing).  Line cut from the point source through the
    middle of the dense clump at $t = 1, 3, 5, 15$ Myr.  of the
    average neutral fraction (top) and temperature (bottom) of the
    clump.}
\end{figure}

\begin{figure}
  \plotone{f12}
  \caption{\label{fig:test3_2} Test 3. (I-front trapping in a dense
    clump and shadowing).  Evolution of the average ionised fraction
    (top) and temperature (bottom) of the overdense clump.}
\end{figure}

\begin{figure}
  \plotone{test3_pdf}
  \caption{\label{fig:test3_pdf} Test 3. (I-front trapping in a dense
    clump and shadowing). Probability distribution function for
    ionized fraction (left) and temperature (right) at 1 Myr (solid),
    3 Myr (dashed), and 15 Myr (dotted) within the dense clump.}
\end{figure}

@@

The problem for this test is contained in a 6.6 kpc box with an
ambient medium of $n_H = 2 \times 10^{-4}\;\cubecm$ and $T = 8000$ K.
The clump is in pressure equilibrium with $n_H = 0.04\; \cubecm$ and
$T = 40$ K.  It has a radius of $r_c = 0.8$ kpc and is centred at
$(x,y,z) = (5, 3.3, 3.3)$ kpc.  The ionized fraction of the entire
domain is zero.  In RT06, the test considered a plane parallel
radiation field with a flux $F_0 = 10^6$ ph s$^{-1}$ cm$^{-2}$
originating from the $y=0$ plane.  Our code can only consider point
sources, so we use a single radiation source located in the centre of
the $y=0$ boundary.  The luminosity of $\dot{N}_\gamma = 3 \times
10^{51}$ ph s$^{-1}$ corresponds to the same flux $F_0$ at 5 kpc.  The
location where the ionisation front trapping can be calculated
analytically \citep{Shapiro04}, and with these parameters, it should
halt at approximately the centre of the clump.  We use the same four
energy groups as in Test 2.

In Fig. \ref{fig:test3_1}, we show neutral fraction and temperature of
a one-dimensional cut through the centre of the dense clump at $z =
0.5$ at $t = 1, 3, 5, 15$ Myr.  The ionisation front is halted at a
little more than halfway through the clump, which is consistent with
the analytical expectation.  The hardness of the $T = 10^5$ K
blackbody spectrum allows the gas outside of the ionisation front to
be photo-heated.  Where the gas is ionised, the temperature is between
10,000 and 20,000 K, but decreases sharply with the ionised fraction.
Fig. \ref{fig:test3_2} depicts the average ionised fraction and
temperature inside the dense clump, which both gradually increase as
the ionisation front propagates through the overdensity.  Our results
are consistent with RT06.  The distribution of ionized fraction and
temperature within the clump are shown in Fig. \ref{fig:test3_pdf} and
agree well with the other codes in RT06.  Finally we show slices of
neutral fraction and temperature in the $z = 0.5$ plane in
Fig. \ref{fig:test3_3}.  The neutral fraction slices prominently show
the sharp shadows created by the clump and demonstrates the
non-diffusivity behavior of ray tracing.  The discretisation of the
sphere creates one neutral cell on the left side of the sphere.  This
inherent artifact to the initial setup carries through the
calculation.  We did not smooth the clump surface like in some of the
RT06 codes, in order to remove this artifact.  It is seen in the
neutral fraction and temperature states at all times and is not a
caused by our ray tracing algorithm.

@@

\begin{figure}
  \plotone{test4_pdf}
  \caption{\label{fig:test4_pdf} Test 4. (Multiple cosmological
    sources). Probability distribution function for neutral fraction
    (left) and temperature (right) at 50 kyr (solid) and 200 kyr
    (dashed).}
\end{figure}

\begin{figure}
  \plotone{f14}
  \caption{\label{fig:test4_1} Test 4. (Multiple cosmological
    sources).  Evolution of the mass- (dashed) and volume-averaged
    (solid) ionised fraction.}
\end{figure}

@@

Fig. \ref{fig:test4_pdf} depicts the distribution of the neutral
fraction and temperature of the entire domain and shows good agreement
with the codes presented in RT06.  We show the growth of the \hii
regions by computing the mass-averaged $x_m$ and volume-averaged $x_v$
ionised fraction in Fig. \ref{fig:test4_1}.  Initially $x_m$ is larger
than $x_v$, and at $t \sim 170$ kyr, the $x_v$ becomes larger.  This
is indicative of inside-out reionisation \citep[e.g.][]{Gnedin00,
  Miralda00, Sokasian03}, where the dense regions around haloes are
ionised first, then the voids are ionised last.  At the end of the
simulation, 65\% of the simulation is ionised.  However by visual
inspection in the slices of electron fraction
(Fig. \ref{fig:test4_2}), there appears to be very good agreement with
C$^2$-ray and FTTE.  By first glance, our result appears to be
different than the RT06 because of the color-mapping.  Our results are
also in good agreement with the multi-frequency version of TRAPHIC
\citep[][see also for better representations of the electron fraction
slices]{Pawlik10}.  In the slices of electron fraction and
temperature, Fig. \ref{fig:test4_2}, the photo-heated regions are
larger than the ionised regions by a factor of 2--3 because of the
hardness of the $T = 10^5$ K blackbody spectrum.

@@

\begin{figure}
  \plotone{test7_pdf}
  \caption{\label{fig:test7_pdf} Test 7. (Photo-evaporation of a dense
    clump).  Probability distribution function for temperature (left)
    and flow Mach number (right) at 10 Myr (solid) and 50 Myr
    (dashed).}
\end{figure}

\begin{figure}
  \plotone{test7_Mach}
  \caption{\label{fig:test7_4} Test 7. (Photo-evaporation of a dense
    clump).  Slices through the clump centre of the flow Mach number
    at t = 10 Myr (left) and 50 Myr (right).}
\end{figure}

@@

Fig. \ref{fig:test7_1} shows cuts of density, temperature, neutral
fraction, and pressure in a line connecting the source and the clump
centre at $t = 1$, 10, and 50 Myr.  At 1 Myr, the ionisation front has
propagated through the left-most 500 pc of the clump.  This heated gas
is now over-pressurised, as seen in the pressure plot in Fig.
\ref{fig:test7_1}, and then expands into the ambient medium.  This
expansion creates a photo-evaporative flow, seen in many star forming
regions \citep[e.g. M16;][]{Hester96} as stars irradiate nearby cold,
dense overdensities.  These flows become evident in the density at 10
Myr, seen both in the line cuts and slices (Fig. \ref{fig:test7_2}).
They have temperatures up to 50,000 K.  At this time, the front has
progressed about halfway through the clump, if one inspects the
neutral fraction.  However the high energy photons have heated all but
the rear surface of the clump.  At the end of the test ($t = 50$ Myr),
only the core and its associated shadow is neutral, seen in Fig.
\ref{fig:test7_3}.  The core has been compressed by the surrounding
warm medium, thus causing the higher densities seen at $t = 50$ Myr.
The non-spherical artifacts on the inner boundary of the warm
outermost shell are caused by the initial discretisation of the
sphere, as discussed in \S\ref{sec:test3}.  Next
Fig. \ref{fig:test7_pdf} shows the distributions of temperature and
flow Mach number in the entire domain at 10 Myr and 50 Myr, showing
similar behavior as the codes in RT09.  Lastly in
Fig. \ref{fig:test7_4}, we show slices of the flow Mach number at 10
Myr and 50 Myr, showing the supersonic photo-evaporative flows that
originate from the clump.

@@

We thank Ji-hoon Kim, Richard Klein, and Jeff Oishi for helpful
comments on the manuscript.  J.~H.~W. is supported by NASA through
Hubble Fellowship grant \#120-6370 awarded by the Space Telescope
Science Institute, which is operated by the Association of
Universities for Research in Astronomy, Inc., for NASA, under contract
NAS 5-26555.  Computational resources were provided by NASA/NCCS award
SMD-09-1439.  The majority of the analysis and plots were done with
\texttt{yt} \citep{yt_full_paper}.
